Understanding the Impact of Socioeconomic and Regional Factors on Health and Educational Outcomes

### 2024-12-08 analysis.qmd
reformatted qmd

---
title: "analysis"
format: html
editor: visual
---

## Load Data & Create Factors

```{r, echo= FALSE, warning = FALSE}
library("dplyr")
library(mice)
library(sjlabelled)
library(tidyverse)

combined <- read.csv('https://github.com/wujenny214/F24_STATS_FINAL/raw/refs/heads/main/combined.csv')
```

```{r}
head(combined)
```

```{r}
combined <- combined |> mutate(State_fac = factor(combined$State), Region_fac = factor(combined$Region, levels=c('South', 'West', 'Northeast', 'Midwest')))
```

```{r}
combined <- combined %>% select(-X, -Spending.per.Pupil)
```

```{r}
head(combined)
```

## Imputation

```{r}
sub <- combined |> select(c("Life.Expectancy","Residential.Segregation.Index","X..Limited.Access.to.Healthy.Foods","X..Uninsured.Adults","Median.Household.Income", "School.Funding.Adequacy", "Median.Prop.Tax", "Region_fac"))
sub <- unlabel(sub) #unlabel the data (labels cause problem for the mice function)
sub.imp <- mice(sub, m=5, method="pmm", print=FALSE)
```

```{r}
#Need to convert school funding adequacy imputed values to school funding categorical variable before using in models
# Imputated data has 5 dataframes within a list, need to run a loop for all five datasets to create a categorical variable for each set.

@@

# Perform the imputation
sub$School.Funding.Cat <- ifelse(sub$School.Funding.Adequacy > 0, 1, 0)  # Add the new variable to the dataset
sub.imp <- mice(sub, m = 5, method = "pmm", print = FALSE)

@@

```{r}
# To create a School.Funding.Categorical variable, create a function and run the loop

@@

```{r}
long_data <- complete(sub.imp, action = "long", include = TRUE)  # Include original data

# Check the structure
str(long_data)

# Ensure that `School.Funding.Cat` is correctly populated
long_data$School.Funding.Cat <- ifelse(long_data$School.Funding.Adequacy > 0, 1, 0)

if (!".imp" %in% colnames(long_data) || !".id" %in% colnames(long_data)) {
  stop("`.imp` or `.id` columns are missing in the long data!")
}
```

## Q1

```{r}
#MLR
# Outcome = life expectancy #Predictors = residential segregation, limited access to healthy foods, uninsured adults, region_fac as interaction with residential segregation (or any other variable)

mlrmodel <- lm(
  Life.Expectancy ~ Residential.Segregation.Index +
    X..Limited.Access.to.Healthy.Foods +
    X..Uninsured.Adults,
  data = combined)
summary(mlrmodel)

@@

```{r}
mlr.imp.mod <- with(sub.imp, lm(Life.Expectancy ~ Residential.Segregation.Index +
    X..Limited.Access.to.Healthy.Foods +
    X..Uninsured.Adults + 
    Region_fac))

summary(pool(mlr.imp.mod))
```

```{r}
mlr.imp.mod2 <- with(sub.imp, lm(Life.Expectancy ~ Residential.Segregation.Index +
    X..Limited.Access.to.Healthy.Foods +
    X..Uninsured.Adults + 
    Region_fac + 
    Region_fac * Residential.Segregation.Index))

summary(pool(mlr.imp.mod2))
```

\@ UZO: Add plots to check assumptions: QQ plot, nested F-test, write p-value interpretations, VIF, Cooks distance

@@

# Nested F-Test
anova(reduced_model, mlrmodel)
```

@@

# Identify Influential Points
influential <- which(cooksd > 4/(nrow(combined) - length(coef(mlrmodel))))
print(influential)
```

@@

## Q2

@@

library("modelsummary")
modelsummary(lmod,
             fmt = fmt_significant(2),
             shape = term ~ model + statistic,
             statistic = c("std.error","conf.int","p.value"),
             exponentiate = FALSE,
             gof_map=NA)

@@

```{r}
imp_mods <- with(sub_imp, glm(School.Funding.Cat ~ Residential.Segregation.Index + Median.Household.Income + Median.Prop.Tax), family="binomial")

modelsummary(pool(imp_mods),
             fmt = fmt_significant(2),
             shape = term ~ model + statistic,
             statistic = c("std.error","conf.int","p.value"),
             exponentiate = FALSE,
             gof_map=NA)

@@

```{r}
# Creating the AUC curve

@@

set.seed(42)
train_idx <- sample(1:nrow(long_data), size = 0.7 * nrow(long_data))
train_data <- long_data[train_idx, ]
test_data <- long_data[-train_idx, ]

glm_model <- glm(School.Funding.Cat ~ Residential.Segregation.Index + Median.Household.Income + Median.Prop.Tax, data = train_data, family = binomial)

# Step 4: Make predictions on the test set
test_probs <- predict(glm_model, newdata = test_data, type = "response")
test_pred <- ifelse(test_probs > 0.5, 1, 0)  # Classify based on threshold 0.5

# Step 5: Create a confusion matrix
conf_matrix <- confusionMatrix(factor(test_pred), factor(test_data$School.Funding.Cat))
print(conf_matrix)
```

>>>>

---
title: "Understanding the Impact of Socioeconomic and Regional Factors on Health and Educational Outcomes"
author: "Rishika Randev, Jenny Wu, Uzoma Uwazurike Jr., Shiyue Zhou"
instructor: "Andrea Lane"
institution: "Duke University, IDS 702 Data Modeling and Presentation"
format: html
editor: visual
execute:
  echo: false
  warning: false
---

## **Abstract**

## **Introduction**

### Background

### Research Question

### Data Overview

## **Methods**

### Variable Selection

### Exploratory Data Analysis & Data Preprocessing

### Model Fitting & Assessment

```{r, echo= FALSE, warning = FALSE, message=FALSE}
library("dplyr")
library(mice)
library(sjlabelled)
library(tidyverse)
library("modelsummary")
```

```{r,echo= FALSE, warning = FALSE, message=FALSE}
combined <- read.csv('https://github.com/wujenny214/F24_STATS_FINAL/raw/refs/heads/main/combined.csv')
```

```{r, echo= FALSE, warning = FALSE}
# Making the state_fac variable a factor variable for future use
combined <- combined |> mutate(State_fac = factor(combined$State), Region_fac = factor(combined$Region, levels=c('South', 'West', 'Northeast', 'Midwest')))
```

```{r, echo= FALSE, warning = FALSE}
## Imputation Code 
# In this block, we are looking to use predictive mean matching to combing regression methods nad hot deck imputing to predicted value closest to the predicted value of the observation with missing data we also needed to convert school funding adequacy imputed values to school funding categorical variable before using in models

@@

# Add School.Funding.Cat to the orginal dataset. Since we do not want to imputate on the categorical outcome variable,  this creation of a variable exists to create the proper dataframe within the imputated data 
sub$School.Funding.Cat <- ifelse(sub$School.Funding.Adequacy > 0, 1, 0)

# Perform the imputation
sub.imp <- mice(sub, m = 5, method = "pmm", print = FALSE)

@@

```{r, echo= FALSE, warning = FALSE}
# Since we should not imputate on the outcome, but need to imputate the School.Funding.Adequacy variable, we are feeding the School.Funding.Adequecy variable into a ifelse conditional that will replaced the impuated outcome variable.

@@

```{r, echo= FALSE, warning = FALSE, include = FALSE }
# To have a full dataset, we are taking the first imputated dataset R provided and using that as our full dataset sample. 
completed_data <- complete(sub.imp, 1)

# Check the structure
str(completed_data)

# Ensure that there are no missing data 
colSums(is.na(completed_data))

@@

## Q1 Life Expectancy; MLR

### **Research Question 1: Life Expectancy**

### Model Fitting

```{r, echo= FALSE, warning = FALSE}
# MLR
# Outcome = life expectancy 
# Predictors = residential segregation, limited access to healthy foods, uninsured adults, region_fac as interaction with residential segregation (or any other variable)

# To get a general sense of what the data tells us, we use the orignial dataset with missing values.

mlrmodel <- lm(
  Life.Expectancy ~ Residential.Segregation.Index +
    X..Limited.Access.to.Healthy.Foods +
    X..Uninsured.Adults,
  data = combined)

# Renaming the output names 
renamed <- c(
  "Residential.Segregation.Index" = "Residential Segregation Index (RSI)",
  "X..Limited.Access.to.Healthy.Foods" = "Access to Healthy Foods",
  "X..Uninsured.Adults" = "Pct. of Uninsured Adults")

modelsummary(mlrmodel, fmt = fmt_significant(2),
             coef_rename = renamed, 
             shape = term ~ model + statistic,
             statistic = c("std.error","conf.int","p.value"),
             exponentiate = FALSE,
             gof_map=NA,
             )

### Regression Results

```{r, echo= FALSE, warning = FALSE}
# Running a regression with the imputated data to understand the impact of missing data

mlr.imp.mod <- with(sub.imp, lm(Life.Expectancy ~ Residential.Segregation.Index +
    X..Limited.Access.to.Healthy.Foods +
    X..Uninsured.Adults + 
    Region_fac))

renamed <- c(
  "Residential.Segregation.Index" = "Residential Segregation Index (RSI)",
  "X..Limited.Access.to.Healthy.Foods" = "Access to Healthy Foods",
  "X..Uninsured.Adults" = "Pct. of Uninsured Adults",
  "Region_facWest" = "Region.W", 
  "Region_facNortheast" = "Region.NE",
  "Region_facMidwest" = "Region.MidW"
  )

modelsummary(mlr.imp.mod, fmt = fmt_significant(2),
             coef_rename = renamed, 
             shape = term ~ model + statistic,
             statistic = c("std.error","conf.int","p.value"),
             exponentiate = FALSE,
             gof_map=NA,
             )
```

```{r echo= FALSE, warning = FALSE}
# Running a regression with the interaction term between residential segregation on the regions

mlr.imp.mod2 <- with(sub.imp, lm(Life.Expectancy ~ Residential.Segregation.Index +
    X..Limited.Access.to.Healthy.Foods +
    X..Uninsured.Adults + 
    Region_fac + 
    Region_fac * Residential.Segregation.Index))

renamed <- c(
  "Residential.Segregation.Index" = "Residential Segregation Index (RSI)",
  "X..Limited.Access.to.Healthy.Foods" = "Access to Healthy Foods",
  "X..Uninsured.Adults" = "Pct. of Uninsured Adults",
  "Region_facWest" = "Region.W", 
  "Region_facNortheast" = "Region.NE",
  "Region_facMidwest" = "Region.MidW", 
  "Residential.Segregation.Index:Region_facWest" = "RSI * Region.W ", 
  "Residential.Segregation.Index:Region_facNortheast" = "RSI * Region.NE",
  "Residential.Segregation.Index:Region_facMidwest" = "RSI * Region.MidW"
  )

modelsummary(mlr.imp.mod2, fmt = fmt_significant(2),
             coef_rename = renamed, 
             shape = term ~ model + statistic,
             statistic = c("std.error","conf.int","p.value"),
             exponentiate = FALSE,
             gof_map=NA,
             )

### Evaluation & Assessment of Model

@@

# Nested F-Test
anova(reduced_model, mlrmodel)

@@

```{r include = FALSE}
# Identify Influential Points
influential <- which(cooksd > 4/(nrow(combined) - length(coef(mlrmodel))))
print(influential)
```

@@

## Q2 **School Funding Adequacy; GLM**

### **Research Question 2: School Funding Adequacy**

### Model Fitting

@@

library("modelsummary")

#log-odds scale
modelsummary(lmod,
             fmt = fmt_significant(2),
             shape = term ~ model + statistic,
             statistic = c("std.error","conf.int","p.value"),
             exponentiate = FALSE,
             gof_map=NA)

@@

### Regression Results

```{r}
imp_mods <- with(sub.imp, glm(School.Funding.Cat ~ Residential.Segregation.Index + Median.Household.Income + Median.Prop.Tax), family="binomial")

#log-odds scale
modelsummary(pool(imp_mods),
             fmt = fmt_significant(2),
             shape = term ~ model + statistic,
             statistic = c("std.error","conf.int","p.value"),
             exponentiate = FALSE,
             gof_map=NA)

@@

### Evaluation and Assessment of Model

```{r}
# Creating the confusion matrix

@@

# set.seed(42)
# train_idx <- sample(1:nrow(long_data), size = 0.5 * nrow(long_data))
# train_data <- long_data[train_idx, ]
# test_data <- long_data[-train_idx, ]

t_glm_model <- glm(School.Funding.Cat ~ Residential.Segregation.Index + Median.Household.Income + Median.Prop.Tax, data = completed_data, family = binomial)

# Make predictions on the test set
test_probs <- predict(t_glm_model, newdata = completed_data, type = "response")
test_pred <- ifelse(test_probs > 0.5, 1, 0)  # Classify based on threshold 0.5
```

```{r}
# Create a confusion matrix
library("caret")
conf_matrix <- confusionMatrix(factor(test_pred), factor(completed_data$School.Funding.Cat))
#print(conf_matrix)

library(pander)

# Format and display confusion matrix as a Markdown table
pander(conf_matrix, caption = "Confusion Matrix")

```{r}
#Creating the AUC curves
roc_result <- roc(
  completed_data$School.Funding.Cat,
  predict(t_glm_model, type = "response"),
  legacy.axes = TRUE,
  plot = TRUE,
  print.thres = 0.5,
  print.auc = FALSE  # Disable automatic AUC
)

# Add AUC manually
text(
  x = 1.2,  # Adjust X-coordinate
  y = 1,  # Adjust Y-coordinate
  labels = sprintf("AUC = %.3f", auc(roc_result)),  # Add AUC value
  cex = 1,  # Label size
  col = "red"  # Label color
)
```

```{r}
